### 2016-05-30 text/Chapter09.tex
Edits and tweaks (Chapter 8 and 9 mostly)

Some of the most fundamental academic work is not the one solving hard research problems, but the
one that changes how we understand the world. Some philosophers argue that \emph{language} is the
key for understanding how we think, while in science the dominant thinking is determined by
\emph{paradigms} \cite{philosophy-kuhn} or \emph{research programmes} \cite{philosophy-lakatos}.
In a way, programming languages play a similar role for computer science and software development.

This thesis aims to change how developers and programming language designers think about
\emph{context} or \emph{execution environments} of programs. Such context or execution environment
has numerous forms -- mobile applications access network, GPS location or user's personal data;
journalists obtain information published on the web or through open government data initiatives;
in dataflow programming, we have access to past values of expressions.
This thesis aims to change our understanding of \emph{context} so that the above examples are
viewed uniformly through a single programming language abstraction, which we call \emph{coeffects},
rather than as disjoint cases.

In this chapter, we give a brief overview of the technical contributions of this thesis
(Section~\ref{sec:conc-summary}). The thesis looks at two kinds of context-dependence, identifies
common patterns and captures those using two \emph{coeffect calculi}. It gives formal semantics
using categorical foundations and uses it as a basis for prototype implementation. Finally,
Section~\ref{sec:conc-conclusions} concludes the thesis.

@@

The main contribution of the chapter is that it presents the disjoint language features in a
unified way. We show type systems and semantics for many of the languages, illuminating
the fact that they are closely related.

@@

We define a type system for the calculus (Section~\ref{sec:flat-calculus}),
discuss how to resolve the ambiguity inherent in context-aware programming for sample
languages we consider (Section~\ref{sec:flat-unique}) and study equational properties of the
calculus (Section~\ref{sec:flat-syntax}). In the flat coeffect calculus, $\beta$-reduction and
$\eta$-expansion do not generally preserve the type of an expression, but we identify two
conditions when this is the case.

@@

We unify systems with \emph{per-variable} contextual properties in Chapter~\ref{ch:structural},
which describes our \emph{structural coeffect calculus} (Section~\ref{sec:struct-calculus}).
Similarly to the flat variant, the calculus is parameterized by a \emph{structural coeffect algebra}.
Concrete instances of the calculus track bounded variable use (\ie~how many times is a variable
accessed), dataflow properties (how many past values are needed) and liveness (\ie~can variable
be accessed).

The structural coeffect calculus has desirable equational properties (Section~\ref{sec:struct-syntax}).
In particular, type preservation for $\beta$-reduction and $\eta$-expansion holds for all instances of the
structural coeffect calculus. This follows from the fact that structural coeffect
associates contextual requirements with individual variables and preserves the connection by
including explicit structural rules (weakening, exchange and contraction).

@@

Both of the coeffect calculi are parameterized and can be instantiated to obtain a language with
a type system that tracks concrete notion of context dependence. The coeffect calculus can be seen
as a framework for building concrete domain-specific context-aware programming languages. In
addition to the type system, the framework also provides a way for defining the semantics of
concrete domain-specific context-aware languages, guides their implementation and simplifies
safety proofs.

@@

A part of the thesis is an interactive web-based essay (\url{http://tomasp.net/coeffects}),
which implements three simple context-aware languages. In addition to more traditional prototype
language implementations, the essay provides a number of novel features. As usual, it lets users
write and evaluate sample programs, but it also lets them explore the details of the coeffect
theory -- this includes an interactive way of exploring typing derivations and program semantics.
The essay also provides a number of case studies that highlight interesting aspects of the theory.

The implementation (Chapter~\ref{ch:impl}) serves two purposes. First, it links together all parts
of the theory developed in this thesis (Section~\ref{sec:impl-case}). It uses the two coeffect calculi
to build a language framework that is then instantiated to implement three concrete context-aware
languages. The  framework provides the type system and translation (based on the
comonadically-inspired semantics) that is used for program execution. This shows the practical
benefits of using the theory of coeffects as a basis of real-world programming languages.

Second, the implementation uses novel kind of medium (Section~\ref{sec:impl-essay}) to make our work
\emph{accessible} and \emph{explorable}. Building a production-ready programming language is outside
the scope of the  thesis, so instead, our goal is to explain the theory and inspire authors of
real-world programming languages to include support for context-aware programming, ideally using
coeffects as a sound foundation. For this reason, we make the implementation accessible over web
without requiring installation of specialized software and we make it explorable to encourage
active reading.

>>>>

Some of the most fundamental academic work is not that which solves hard research problems, but
that which changes how we understand the world. Some philosophers argue that \emph{language} is the
key for understanding how we think, while in science the dominant thinking is determined by
\emph{paradigms} \cite{philosophy-kuhn} or \emph{research programmes} \cite{philosophy-lakatos}.
Programming languages play a similar role for computer science and software development.

This thesis aims to change how developers and programming language designers think about
\emph{contexts} or \emph{execution environments} of programs. Such contexts or execution environments
have numerous forms -- mobile applications access network, GPS location or user's personal data;
journalists obtain information published on the web or through open government data initiatives;
in dataflow programming, we have access to past values of expressions.
This thesis aims to change our understanding of \emph{context} so that the above examples are
viewed uniformly through a single programming language abstraction, which we call \emph{coeffects},
rather than as disjoint cases.

In this chapter, we give a brief overview of the technical contributions of this thesis
(Section~\ref{sec:conc-summary}). The thesis looks at two kinds of context-dependence, identifies
common patterns and captures those using two \emph{coeffect calculi}. It gives a formal semantics
using categorical insights and uses it as a basis for prototype implementation. Finally,
Section~\ref{sec:conc-conclusions} concludes the thesis.

@@

The main contribution of Chapter~\ref{ch:applications} is that it presents the disjoint language
features in a unified way. We show type systems and semantics for many of the languages,
illuminating the fact that they are closely related.

@@

We define a type system for the calculus (Section~\ref{sec:flat-calculus}),
discuss how to resolve the ambiguity inherent in context-aware programming for sample
languages we consider (Section~\ref{sec:flat-unique}) and study equational properties of the
calculus (Section~\ref{sec:flat-syntax}). In the flat coeffect calculus, $\beta$-reduction and
$\eta$-expansion do not generally preserve the type of an expression, but we identify two
conditions satisfied by two of our instances when this is the case.

@@

We characterise systems with \emph{per-variable} contextual properties in Chapter~\ref{ch:structural},
which describes our \emph{structural coeffect calculus} (Section~\ref{sec:struct-calculus}).
Similarly to the flat variant, the calculus is parameterized by a \emph{structural coeffect algebra}.
Concrete instances of the calculus track bounded variable use (\ie~how many times is a variable
accessed), dataflow properties (how many past values are needed) and liveness (\ie~can variable
be accessed).

The structural coeffect calculus has desirable equational properties (Section~\ref{sec:struct-syntax}).
In particular, type preservation for $\beta$-reduction and $\eta$-expansion holds for all instances of the
structural coeffect calculus. This follows from the fact that structural coeffects
associate contextual requirements with individual variables and preserves the connection by
including explicit structural rules (weakening, exchange and contraction).

@@

Both forms of coeffect calculi are parameterized and can be instantiated to obtain a language with
a type system that tracks concrete notion of context dependence. The coeffect calculus can be seen
as a framework for building concrete domain-specific context-aware programming languages. In
addition to the type system, the framework also provides a way for defining the semantics of
concrete domain-specific context-aware languages, guides their implementation and simplifies
safety proofs.

@@

Associated with this essay is an interactive web-based essay, which implements three simple
context-aware languages (Chapter~\ref{ch:impl}). In addition to traditional prototype language
implementations, the essay provides a number of novel features. As usual, it lets users
write and evaluate sample programs, but it also lets them explore the details of the coeffect
theory -- this includes an interactive way of exploring typing derivations and program semantics.
The essay also provides a number of case studies that highlight interesting aspects of the theory.

The implementation (Section~\ref{ch:impl-theory}) serves two purposes. First, it links together all parts
of the theory developed in this thesis (Section~\ref{sec:impl-case}). It uses the flat and structural
coeffect calculi to build a language framework that is then instantiated to implement three concrete
context-aware languages. The  framework provides the type system and translation (based on the
comonadically-inspired semantics) that is used for program execution. This shows the practical
benefits of using the theory of coeffects as a basis of real-world programming languages.

Second, web-based essay (\url{http://tomasp.net/coeffects}) discussed in Section~\ref{sec:impl-essay}
is novel from the pedagogical perspective. It makes our work \emph{accessible} and \emph{explorable}.
Building a production-ready programming language is outside the scope of the  thesis, so instead,
our goal is to explain the theory and inspire authors of real-world programming languages to
include support for context-aware programming, ideally using coeffects as a sound foundation.
For this reason, we make the implementation accessible over web without requiring installation
of specialized software and we make it explorable to encourage \emph{active reading}.
